Optimal Algorithms for Dynamic Balanced Graph Partitioning

### 2020-05-15 paper/disc20clustering.tex
Finalizations

	  Next, we argue that a feasible partition exists when the process terminates.
	 This implies that a feasible partition exists after any earlier step as well.
%	 Assume for contradiction that the partition is not feasible.	 
	 Since there are at least $\ell+1$ isolated nodes left,
	 there must be two isolated nodes $x^*$ and $y^*$,
	  with the same cluster of origin,
	 i.e., $I(\{x^*\}) = I(\{y^*\})$.
	 Consider the partition $P^*$
	  obtained from the initial partition after swapping $x_0$ and $y_0$
	 with $x^*$ and $y^*$ (respectively).
	 In this partition,
	 the ground set $\{x_0,y_0\}$ is collocated in the cluster $I(\{x^*,y^*\})$.
	 Note that after the first request $\{x_0,y_0\}$,
	   we issue requests only between nodes that have the same cluster of origin and
	   all these nodes are collocated in $P^*$.
	 Therefore all ground sets constructed so far are collocated in  $P^*$,
	 i.e., $P^*$ is a feasible partition.
%	 Hence, there are at most $\ell$ isolated nodes after the termination.

@@

	\ALG performs at least one swap at each step $i$,
	and some ground set grows.
%	In other words,
%	some ground set grows by one node at each of these steps.
	Consider any ground set $C^* \neq B$ after the termination.
	This ground set has grown exactly $|C^*|-1$ times until the the termination. 
	Let $\mathcal{S}$ be the set of all ground sets after the process terminates.
	Thus, $\mathcal{S}$ includes ground sets $B$,
	$\{x_0,y_0\}$,
	and (up to) $\ell+2$ singleton ground sets.
	Among the remaining ground sets in $\mathcal{S}$,
	no two ground sets have the same origin.
	Otherwise,
   the smaller ground set is either a singleton,
   which contradicts the bound $\ell+2$ on the number of singletons,
	or we have joined nodes to it at some step,
	contradicting our choice of the largest $C_i$ at step $i$.
	Hence,  there are at most $\ell-1$ such ground sets,
	one per possible cluster of origin,
	 excluding the cluster containing $B$. 
	Therefore,  $|\mathcal{S}| \leq 1 + 1 + (\ell+2)  + (\ell-1) = 2\ell+3$.
	Note that among all non-singleton ground sets in $\mathcal{S}$,
	 only $B$ does not grow during the process.
	 Thus,
    the total number of times that 
	a ground set in $\mathcal{S}$  has grown is
	$	\sum_{C^* \in \mathcal{S} } (|C^*|-1) - (k-1) $
	\begin{align*}
	=\sum_{C^* \in \mathcal{S} } |C^*| - 
		\sum_{C^* \in \mathcal{S} } 1 - (k-1) 
	\geq  k  \ell - (2\ell+3) - (k-1) =  (k-2)(\ell-1) - 4,
	\end{align*}
	which bounds the number of swaps performed by \ALG.
	The competitive ratio is then
	$\ALG / \OPT \geq ((k-2)(\ell-1) - 4) / 2$.
	%%
	\begin{comment}
	\ALG performs at least one swap  every time   a~ground set is constructed.
	Each expansion reduces the number of isolated nodes by $1$ or $2$.
	The decrease of two isolated nodes occurs when the largest component on  $I(x_i)$ is still singleton; this occurs at most $\ell$ times.
	We start with $k \cdot \ell - (k-1)$ isolated nodes (a~single $k-1$ ground set is revealed prior to $\{x_0, y_0\}$), and we end with isolated $\ell+1$ nodes.
	Hence, the total number of swaps of \ALG is at least $k\cdot \ell - k - 2 \ell$.
	The competitive ratio is then $\ALG / \OPT \geq (k\cdot \ell - k - 2\ell) / 2$.
	\end{comment}
	%%
\end{proof}

@@

The algorithm in this section and the algorithm for $k=3$ (cf. Section~\ref{sec:k3}) are modified versions of the algorithm DET from~\cite{repartition-disc}.
The difference is in the choice of partition after a~component merge.
In DET, the partition was arbitrary.
In the algorithm for the learning model,
we choose a~component respecting partition closest to the initial partition.
In the algorithm from Section~\ref{sec:k3}, we choose the partition closest to the current partition (a repartition of minimum cost).
%A generic sketch of these two algorithms can be found in the appendix (Algorithm \ref{alg:ppl}).

@@

Now we describe the algorithm \PPL.
On each inter-cluster request $\{u,v\}$, 
\PPL creates a new components by merging
 the two components that contain nodes $u$ and $v$.
In order to collocate nodes of the new component,
\PPL moves to a~component respecting partition that minimizes the distance to the initial partition $P_I$.
The scheme of the algorithm can be found in the  appendix (Algorithm \ref{alg:ppl}).

\PPL never moves to a~partition that is more than $ \Delta^*:= \Delta(P_F) $   migrations away from $P_I$.
This invariant latter ensures us that \PPL does not pay too much while recovering $P_F$.
We emphasize that a~\emph{repartitioning} by \PPL replaces the current partition $P$ with a~perfect partition closest to $P_I$.
This way \PPL never moves to a~partition beyond the distance $\Delta^*$.

@@

\stefan{I think we can delete this paragraph: now we discussed everything
in the individual sections, no?}
Although we have described our algorithms globally so far, we note that these allow for efficient distributed implementations. 
The algorithm PPL from Section~\ref{sec:ppl} can be distributed
similarly to the approach in~\cite{sigmetrics19_partitioning}.
The algorithm for $k=2$ from Section~\ref{sec:k2} performs only local communication for each request: counters are kept on the clusters and updated locally, and each migration is local within two clusters that reached the collocation threshold $\lambda$.
Furthermore, we proposed an efficient distributed implementation of the algorithm for $k=3$ in Section~\ref{sec:k3}.

@@

%\section{Additional Figures}
\begin{comment}
\begin{figure}[H]	\label{fig:timeline}
	\centering
\begin{tikzpicture}
\draw [->] (-10mm,0) -- (5,0);
\draw [dotted] (-13mm,0) -- (-10mm,0);

@@

\draw [] (5mm,2pt) -- (5mm,-2pt)
node [black,midway,below=0pt,xshift=1pt] {$t''$};
%% vertical arrow 
%\draw [->] (3mm,14pt) -- (4.4mm,4pt)
%node [black,midway,above=0pt,xshift=-4pt] {separation of  $\{u,w\}$};
%% vertical arrow 
%\draw [->] (30mm,-24pt) -- (26.5mm,-7pt)
%node [black,midway,below=0pt,xshift=5pt, yshift=-5pt] {separation of  $\{u,v\}$};
\end{tikzpicture}
\caption{
Illustration of the timeline used in the proof of Theorem \ref{thm:k=2}.
Requests in $R_t$  are only to  $\{u,w\}$ and $\{u,v\}$,
which  arrive  during the interval $(t'',t]$.
Similarly,
request in $R_{t'}$  are to $\{u,w\}$ and some other pair, irrelevant to the analysis.
Hence, requests to $\{u,w\}$ are included in at most two such sets,
which are $R_{t'}$ and $R_t$.
This is because their intervals overlap on $(t'',t']$, shown with red thick line.
}
\end{figure}
\end{comment}

>>>>

	  Next, we argue that a feasible partition exists when the process terminates.
	 This implies that a feasible partition exists after any earlier step as well.
%	 Assume for contradiction that the partition is not feasible.	 
	 Since there are at least $\ell+1$ isolated nodes left,
	 there must be two isolated nodes $x^*$ and $y^*$,
	  with the same cluster of origin,
	 i.e., $I(\{x^*\}) = I(\{y^*\})$.
	 Consider the partition $P^*$
	  obtained from the initial partition after swapping $x_0$ and $y_0$
	 with $x^*$ and $y^*$ (respectively).
	 In this partition,
	 the ground set $\{x_0,y_0\}$ is collocated in the cluster $I(\{x^*,y^*\})$.
	 Note that after the first request $\{x_0,y_0\}$,
	   we issue requests only between nodes that have the same cluster of origin and
	   all these nodes are collocated in $P^*$.
	 Therefore all ground sets constructed so far are collocated in  $P^*$,
	 and it is a feasible partition.
%	 Hence, there are at most $\ell$ isolated nodes after the termination.

@@

	\ALG performs at least one swap at each step $i$,
	and some ground set grows.
%	In other words,
%	some ground set grows by one node at each of these steps.
	Consider any ground set $C^* \neq B$ after the termination.
	This ground set has grown exactly $|C^*|-1$ times until the the termination. 
	Let $\mathcal{S}$ be the set of all ground sets after the process terminates.
	Thus, $\mathcal{S}$ includes ground sets $B$,
	$\{x_0,y_0\}$,
	and (up to) $\ell+2$ singleton ground sets.
	Among the remaining ground sets in $\mathcal{S}$,
	no two ground sets have the same origin.
	Otherwise,
   the smaller ground set is either a singleton,
   which contradicts the bound $\ell+2$ on the number of singletons,
	or we have joined nodes to it at some step,
	contradicting our choice of the largest~$C_i$ at step $i$.
	Hence,  there are at most $\ell-1$ such ground sets,
	one per possible cluster of origin,
	 excluding the cluster containing $B$. 
	Therefore,  $|\mathcal{S}| \leq 1 + 1 + (\ell+2)  + (\ell-1) = 2\ell+3$.
	Note that among all non-singleton ground sets in $\mathcal{S}$,
	 only $B$ does not grow during the process.
	 Thus,
    the total number of times that 
	a ground set in $\mathcal{S}$  has grown is
	$	\sum_{C^* \in \mathcal{S} } (|C^*|-1) - (k-1) $
	\begin{align*}
	=\sum_{C^* \in \mathcal{S} } |C^*| - 
		\sum_{C^* \in \mathcal{S} } 1 - (k-1) 
	\geq  k  \ell - (2\ell+3) - (k-1) =  (k-2)(\ell-1) - 4,
	\end{align*}
	which bounds the number of swaps performed by \ALG.
	The competitive ratio is then
	$\ALG / \OPT \geq ((k-2)(\ell-1) - 4) / 2$.
	%%
	\begin{comment}
	\ALG performs at least one swap  every time   a~ground set is constructed.
	Each expansion reduces the number of isolated nodes by $1$ or $2$.
	The decrease of two isolated nodes occurs when the largest component on  $I(x_i)$ is still singleton; this occurs at most $\ell$ times.
	We start with $k \cdot \ell - (k-1)$ isolated nodes (a~single $k-1$ ground set is revealed prior to $\{x_0, y_0\}$), and we end with isolated $\ell+1$ nodes.
	Hence, the total number of swaps of \ALG is at least $k\cdot \ell - k - 2 \ell$.
	The competitive ratio is then $\ALG / \OPT \geq (k\cdot \ell - k - 2\ell) / 2$.
	\end{comment}
	%%
\end{proof}

@@

The algorithm in this section and the algorithm for $k=3$ (cf. Section~\ref{sec:k3}) are modified versions of the algorithm DET from~\cite{repartition-disc}.
The difference is in the choice of partition after a~component merge.
In DET, the partition was arbitrary.
In the algorithm for the learning model,
we choose a~component respecting partition closest to the initial partition.
In the algorithm from Section~\ref{sec:k3}, we choose the partition closest to the current partition (a~repartition of minimum cost).
%A generic sketch of these two algorithms can be found in the appendix (Algorithm \ref{alg:ppl}).

@@

Now we describe the algorithm \PPL.
On each inter-cluster request $\{u,v\}$, 
\PPL creates new components by merging
 the two components that contain nodes $u$ and $v$.
In order to collocate nodes of the new component,
\PPL moves to a~component respecting partition that minimizes the distance to the initial partition $P_I$.
The scheme of the algorithm can be found in the  appendix (Algorithm \ref{alg:ppl}).

\PPL never moves to a~partition that is more than $ \Delta^*:= \Delta(P_F) $   migrations away from~$P_I$.
This invariant latter ensures us that \PPL does not pay too much while recovering $P_F$.
We emphasize that a~\emph{repartitioning} by \PPL replaces the current partition $P$ with a~perfect partition closest to $P_I$.
This way \PPL never moves to a~partition beyond the distance $\Delta^*$.

@@

Our algorithms allow for efficient distributed implementations. 
The algorithm PPL from Section~\ref{sec:ppl} can be distributed
similarly to the approach in~\cite{sigmetrics19_partitioning}.
The algorithm for $k=2$ from Section~\ref{sec:k2} performs only local communication for each request: counters are kept on the clusters and updated locally, and each migration is local within two clusters that reached the collocation threshold $\lambda$.
Furthermore, we proposed an efficient distributed implementation of the algorithm for $k=3$ in Section~\ref{sec:k3}.

@@

%\section{Additional Figures}
%\begin{comment}
\begin{figure}[H]	\label{fig:timeline}
	\centering
\begin{tikzpicture}
\draw [->] (-10mm,0) -- (5,0);
\draw [dotted] (-13mm,0) -- (-10mm,0);

@@

\draw [] (5mm,2pt) -- (5mm,-2pt)
node [black,midway,below=0pt,xshift=1pt] {$t''$};
%% vertical arrow 
%\draw [->] (3mm,14pt) -- (4.4mm,4pt)
%node [black,midway,above=0pt,xshift=-4pt] {separation of  $\{u,w\}$};
%% vertical arrow 
%\draw [->] (30mm,-24pt) -- (26.5mm,-7pt)
%node [black,midway,below=0pt,xshift=5pt, yshift=-5pt] {separation of  $\{u,v\}$};
\end{tikzpicture}
\caption{
Illustration of the timeline used in the proof of Theorem \ref{thm:k=2}.
Requests in $R_t$  are only to  $\{u,w\}$ and $\{u,v\}$,
which  arrive  during the interval $(t'',t]$.
Similarly,
request in $R_{t'}$  are to $\{u,w\}$ and some other pair, irrelevant to the analysis.
Hence, requests to $\{u,w\}$ are included in at most two such sets,
which are $R_{t'}$ and $R_t$.
This is because their intervals overlap on $(t'',t']$, shown with red thick line.
}
\end{figure}
%\end{comment}
